Multiple Comparisons in Toolmark Evidence

### 2023-11-21 LPR/index.qmd
Updates

Let us consider the process of toolmark examination - an examiner is given a tool recovered from a suspect as well as a wire recovered from the scene. The examiner will typically take the tool and use it to cut a piece of metal, known as a blade cut, in order to record the markings of the entire cutting surface simultaneously. This process may be repeated at multiple angles in order to record the effect of different contact surfaces on the resulting striations. The resulting blade cut will be compared to the wire using either visual inspection or perhaps in the future a numerical comparison algorithm similar to those developed for bullet striations [@hare2017automatic] and cartridge cases [@zemmels2023].

@@

![Wire cutter with 1.5 cm cutting surface beveled on both sides](../fig/wire_cutter.png){#fig-wire-cutter alt="Wire cutter with 1.5 cm cutting surface beveled on both sides"}

@@

### Random Match Probability

@@

We need to have an explicit relationship between random match probability and amount of signal (assessed by length, number of features, overall striae depth, etc.). Until we have this kind of relationship, it will be hard to use error rates (even if they were known for toolmarks) from algorithms or black-box studies in practical situations.

>>>>

When an examiner conducts a toolmark assessment of wire cut evidence, they are typically working with a tool recovered from the suspect as well as a wire recovered from the scene. The examiner will typically take the tool and use it to cut a piece of metal, known as a blade cut, in order to record the markings of the entire cutting surface simultaneously. This process may be repeated at multiple angles in order to record the effect of different contact surfaces on the resulting striations. The resulting blade cut will be compared to the wire using either visual inspection or perhaps in the future a numerical comparison algorithm similar to those developed for bullet striations [@hare2017automatic] and cartridge cases [@zemmels2023].

@@

![Wire cutter with 1.5 cm cutting surface beveled on both sides](../fig/wire_cutter.png){#fig-wire-cutter}

@@

### Probability of a False Identification

Let us briefly consider a non-exhaustive list of ways an examiner or algorithm might arrive at a false identification:

-   class characteristics match + decider doesn't recognize class characteristics as such

-   coincidental similarity between two items + decider doesn't see/recognize dissimilarities

-   the items randomly match on identifying characteristics

All of these situations may occur during white box studies (though the final situation is very improbable). As a result, perhaps the best estimate of the probability of a false identification that we can obtain is the false identification rate of one or more white box studies. We have previously published a paper detailing the error rates of different firearms and toolmark studies; to our knowledge there are no similar white box studies for wire cuts. However, if we confine ourselves to error rates for striated materials (bullets, aperture shear, ejectors, screwdrivers) we may be able to get a reasonable estimate for the error rate of wire cuts. We assembled 3 black-box error rate studies of striated toolmarks (bullets and firing pins) [@bajic2020; @mattijssen2021; @best2022]; these studies estimate the false positive rate (e.g. the number of different-source identifications relative to the total number of different-source comparisons) rate to be between 0.07235 [@mattijssen2021, firing pins only] and 0.004545 [@best2022]. While there are a number of problems with these studies (which we have detailed in previous papers, i.e. @inconclusives), our goal here is to use the best estimates we can get of false positive error rates in particular, as the false positive rate is the main driver of the problem with multiple comparisons. If we calculate a pooled false positive rate from all studies, we arrive at a false positive rate of $\frac{20 + 56 + 1}{2842 + 774 + 220} = `r sprintf('%0.02f', 77/(2842+774+220))`$, that is, about 2%.

```{r}
comparisons <- examiner_compare*2
fpr = (20+56+1)/(2842+774+220) 
fprformat = sprintf("%0.04f", fpr)
```

If we start out by considering the first scenario (a single tool with a 1.5 cm blade compared to a single 2mm wire) and completely independent non-overlapping comparisons (7.5 comparisons) with only 2 cutting surfaces, we would effectively have 15 independent comparisons. Using this information, we can calculate the probability of at least one false positive error to be `r sprintf("%0.04f%%", 1-(1-fpr)^comparisons * 100)` as shown in Equation \ref{eq:fpr}. While far more comparisons are likely to be performed, these comparisons may have some dependencies (at which point our formula would need to account for the covariance between successive comparisons). Ultimately, the number of comparisons cannot drop below `r comparisons` without leaving some evidence unexamined, which means this overall false positive rate is the lower bound.

This is the simplest situation of the three different scenarios we have considered, and we still have an unacceptably high rate of at least one false positive conclusion. If the wire size decreases or the blade length increases, the overall chance of at least one false positive error will increase because the number of comparisons increases. If examiners consider cuts made at multiple angles, the same conclusion holds - the chance of at least one false positive error will increase from this point.

\begin{align*}
P(\text{False positive error in one comparison}) &= `r fprformat`\\
P(\text{No false positive error in one comparison}) &= `r sprintf('%0.04f', 1-fpr)`\\
P(\text{no false positive errors in } N \text{ comparisons}) &= `r sprintf('%0.04f', 1-fpr)`^N\\
P(\text{at least one false positive error in } N \text{ comparisons}) &= 1 - `r sprintf('%0.04f', 1-fpr)`^N  \tag{A}\label{eq:fpr}\\
&= `r sprintf("%.04f", 1-(1-fpr)^comparisons)` %>% 
\end{align*}

It then becomes relevant to consider how low the error rate would need to be for striated comparisons in order to make wire cut examination plausible. If we control the overall error rate amid $N$ independent comparisons, our estimated FPR for a single comparison can be calculated by solving Equation \ref{eq:fpr-control} for $r$.

Let $F_n:=\{\text{at least one false positive in }n\text{ comparisons}\}$

\begin{align*}
P(F_1) &= r\\
P(\overline{F_1}) &= 1-r\\
P(F_n) &= \left[1-r\right]^n\\
P(\overline{F_n}) &= 1 - \left[P(\overline{F_1})\right]^n = 1 - \left[1 - r\right]^n\tag{B}\label{eq:fpr-control}\\
r &= 1 - \left[1 - P(F_n)\right]^{1/n}
\end{align*}

We can explore what this means visually by considering different values of $N$ and $r$ to see how the overall false positive error rate behaves, as shown in @fig-control-overall-fpr. The astonishing conclusion is that even at the smallest error rate for striated comparisons, 0.45% [@bajic2020], we have about a 3% chance of at least one false positive conclusion when 10 comparisons are made and about a 37.5% chance of at least one false positive conclusion when 100 comparisons are made! These results suggest that under current estimates of false positive error rates, it is nearly impossible for toolmark comparisons of wires to provide compelling, "beyond a reasonable doubt" levels of evidence.

```{r control-overall-fpr}
#| echo: false
#| message: false
#| fig-cap: Exploration of overall error rate estimates for different values of $r$, the false positive rate for a single comparison. Dotted vertical lines show false positive error rate estimates (from left to right, @bajic2020; @best2022; @mattijssen2021). 
#| label: fig-control-overall-fpr
library(ggplot2)
library(tidyr)
single_rmp = exp(seq(-1000, 0, .1))
N = c(10, 100, 1000, 10000)

res = crossing(single_rmp, N) %>%
  mutate(overall = 1 - (1 - single_rmp)^N)

ggplot(res, aes(x = single_rmp, y = overall, group = factor(N), color = factor(N))) + 
  geom_line() +
  coord_cartesian(xlim = c(0, 0.07), ylim = c(0, 1)) + 
  ylab("Probability of at least one false positive\nin N comparisons") + 
  xlab("Probability of false positive for 1 comparison") + 
  ggtitle("Wire Comparison: Controlling Overall False Positive Rate") + 
  geom_vline(xintercept = 0.02, linetype = "dashed", color = "grey40") + 
  geom_vline(xintercept = c(0.00704, 0.0724, 0.004545), linetype = "dotted", color = "grey50") + 
  annotate(geom = "text", x = 0.021, y = 0.125, label = "Striated toolmark error rate estimate\nused in this paper", color = "grey40", hjust=0) + 
  scale_color_discrete("# comparisons") + 
  theme_bw()

```

@@

We need to have an explicit relationship between false positive rate and amount of signal (assessed by length, number of features, overall striae depth, etc.). Until we have this kind of relationship, it will be hard to use error rates (even if they were known for toolmarks) from algorithms or black-box studies in practical situations.
